Tutorial: Quantifying Spatially Resolved Hydration Thermodynamics Using Grid Inhomogeneous Solvation Theory [Article v\versionnumber]

### 2025-05-16 manuscript/manuscript.tex
Added signposting
1) Direct  links to cpptraj, amber manual
2) More explanation for bound command

Furthermore, a recent Python version (3.6 or newer) should be available with the following packages installed:
\begin{itemize}
	\item \software{gisttools} (\url{https://github.com/liedllab/gisttools}, $\geq$ 0.2)
	\item \software{mdtraj} (\url{https://www.mdtraj.org/}, $\geq$ 1.9.7) \cite{McGibbon2015-mdtraj}
	\item \software{matplotlib} (\url{https://matplotlib.org/}, $\geq$ 3.7.0) \cite{Hunter2007-matplotlib}
	\item \software{numpy} (\url{https://numpy.org/}, $\geq$ 1.23.5) \cite{harris2020-numpy}
	\item \software{pandas} (\url{https://pandas.pydata.org/}, $\geq$ 1.5.3) \cite{pandas-2023_8239932}
\end{itemize}
If the reader prefers to skip the MD and GIST calculation, we also provide the GIST output files with this tutorial, such that the post-processing can be done without any expensive calculations.
We additionally provide a Jupyter notebook \cite{Kluyver2016-jupyter,Granger2021-jupyter} with the presented analyses. 
For the visualization of three-dimensional contributions we require molecular visualization software that can handle \software{OpenDX} files, such as \software{PyMol} \cite{pymol} or \software{VMD} \cite{vmd}.
Most of the visualisations and results in this tutorial can also be obtained using the \software{gistpp} program, which is available from \url{https://github.com/KurtzmanLab/Gist-Post-Processing} (code and documentation). \software{Gistpp} can directly manipulate \software{OpenDX} files from the GIST output. 
Its usage and some examples were shown in a previous tutorial on GIST \cite{Ramsey2016}.

\pagebreak %TODO: pagebreak, maybe remove these by better formatting

@@

Using \inlinecode{cpptraj's } \inlinecode{bounds} command, boundaries can be output to a file. 
For example, here is the command for a boundary of \SI{10}{\angstrom} around the ligand LIG:

\begin{lstlisting}[style=cpptraj]
bounds :LIG dx 0.5 offset 20 name Grid out bounds.dat
\end{lstlisting}
\begin{minipage}{\linewidth} %TODO: maybe find a better way to keep listing on same page?
Make sure that the \inlinecode{bounds} command is supplied with the exact same trajectory as in the GIST calculation. 
Any preprocessing steps need to match between the \inlinecode{bounds} command and the \inlinecode{gist} calculation. 
A helper script called \inlinecode{findcentroid.py} for the same purpose is supplied with this tutorial, and can be used as follows:
~
\begin{lstlisting}[style=bash]
python findcentroid.py -i structure.pdb
\end{lstlisting}
\end{minipage}
\paragraph{Run GIST using cpptraj}
GIST is implemented in \software{cpptraj} which requires the topology and the trajectory of the MD simulation to be analyzed.
The command needed to run GIST in cpptraj can be written in an input file (e.g gist.in) for convenience.
The basic command, assuming a grid centered at coordinates (x=\SI{10.5}{\angstrom}, y=\SI{20.1}{\angstrom}, z=\SI{30.1}{\angstrom}) and a grid size of $30\times30\times30$ voxels with \SI{0.5}{\angstrom} spacing, is:

@@

\subsection{Tutorial data}
We will use the crystal structure of streptavidin (PDB code: 1STP \cite{Weber1989-streptavidin-structure}) to start our calculations.
For section \ref{subsec:gist_biotin}, we will use the biotin ligand extracted from the crystal structure.
For the second part in section \ref{subsec:gist_complex}, we will use the whole crystal structure.
For the sake of simplicity, you can download prepared and solvated Amber topologies and structures based on 1STP by running
\begin{lstlisting}[style=bash]
git clone git@github.com:liedllab/gist-tutorial.git
\end{lstlisting}
This will download the tutorial files, including both the manuscript and the examples in the \inlinecode{code} folder.

\subsection{Running GIST for Biotin}
\label{subsec:gist_biotin}
In this example, we will be running a simple GIST calculation of the biotin solute molecule to get familiar with the usage of GIST in \software{cpptraj}.
\subsubsection{System preparation}
First, extract biotin from the crystal structure and generate a topology and coordinate file compatible with Amber or a simulation package of your choice.
The solute molecule should be solvated with a sufficiently large solvent box. 
For periodic boundary conditions, the size of the solvent box should be large enough for the solvent to behave bulk-like so as to reduce artifacts from solvent-mediated interactions between the periodic images. 
This corresponds to a distance between solute and box boundary of at least \SI{12}{\angstrom} for water \cite{Chen2021,Waibl2022-gist-solvents} though we recommend \SI{15}{\angstrom} or higher.
Carefully check the employed force field, protonation state and partial charges.
%a PDB file containing only biotin. Prepare your structure by adding hydrogens, making sure to choose the correct protonation state. Parametrize biotin with the forcefield of your choice. Solvate the parametrized biotin, making sure to choose a water box of large enough size so that the water on the box edges behaves bulk-like (a wall distance of at least 15\AA is recommended).
A good starting point for the parametrization and solvation of small molecules is the Amber tutorial 4b (\url{http://ambermd.org/tutorials/basic/tutorial4b/}).

@@

To decide on a grid size, find the extent (minimum and maximum) of the x-, \mbox{y-,} and z-coordinates of the molecule.
To get bulk-like water properties at the grid border, a buffer distance of \SI{15}{\angstrom} is reasonable for biotin.
The grid parameters can be directly calculated using \software{cpptraj's} \inlinecode{bound} command:
\begin{lstlisting}[style=cpptraj]
parm solvated.parm7
trajin md-01.nc 1 1 1
autoimage !(:WAT) origin
bounds !(:WAT) dx 0.5 offset 30 name MyGrid out bounds.dat
run
\end{lstlisting}
In the above snippet, we calculate the bounds around the whole protein for a grid with \SI{0.5}{\angstrom} side-length voxels and a buffer zone of \SI{15.0}{\angstrom}. 
Note that the offset is given in terms of number of additional voxels, not length dimension. By varying the atom mask in the \inlinecode{bound} command, it is possible to center the grid on a region of interest, similar to the \software{findcentroid.py} script.

>>>>

Furthermore, a recent Python version (3.6 or newer) should be available with the following packages installed:
\begin{itemize}
	\item \software{gisttools} (\url{https://github.com/liedllab/gisttools}, $\geq$ 0.2)
	\item \software{mdtraj} (\url{https://www.mdtraj.org/}, $\geq$ 1.9.7) \cite{McGibbon2015-mdtraj}
	\item \software{matplotlib} (\url{https://matplotlib.org/}, $\geq$ 3.7.0) \cite{Hunter2007-matplotlib}
	\item \software{numpy} (\url{https://numpy.org/}, $\geq$ 1.23.5) \cite{harris2020-numpy}
	\item \software{pandas} (\url{https://pandas.pydata.org/}, $\geq$ 1.5.3) \cite{pandas-2023_8239932}
\end{itemize}
If the reader prefers to skip the MD and GIST calculation, we also provide the GIST output files with this tutorial, such that the post-processing can be done without any expensive calculations.
We additionally provide a Jupyter notebook \cite{Kluyver2016-jupyter,Granger2021-jupyter} with the presented analyses. 
For the visualization of three-dimensional contributions we require molecular visualization software that can handle \software{OpenDX} files, such as \software{PyMol} \cite{pymol} or \software{VMD} \cite{vmd}.
Most of the visualisations and results in this tutorial can also be obtained using the \software{gistpp} program, which is available from \url{https://github.com/KurtzmanLab/Gist-Post-Processing} (code and documentation). \software{Gistpp} can directly manipulate \software{OpenDX} files from the GIST output. 
Its usage and some examples were shown in a previous tutorial on GIST \cite{Ramsey2016}.
For help with the listed cpptraj commands, AMBER atom mask syntax and other AMBER-specific commands, we recommend the following webresources:\\
\begin{itemize}
\item The website https://amberhub.chpc.utah.edu/ which is managed by the Cheatham III group, 
\item The cpptraj manual at https://raw.githubusercontent.com/Amber-MD/cpptraj/master/doc/CpptrajManual.pdf 
\item and the AMBER manual \cite{amber24} at https://ambermd.org/Manuals.php.
\end{itemize}
\pagebreak %TODO: pagebreak, maybe remove these by better formatting

@@

Using \inlinecode{cpptraj's } \inlinecode{bounds} command, boundaries can be output to a file. 
For example, here is the command for a boundary of \SI{15}{\angstrom} around the ligand LIG:
\begin{lstlisting}[style=cpptraj]
bounds :LIG dx 0.5 offset 30 name Grid out bounds.dat
\end{lstlisting}
The atom mask \inlinecode{:LIG} specifies the residue named LIG to build the boundaries around, with dx and offset providing the actual dimensions of the boundaries, i.e. an offset of 30 times 0.5\AA.
\begin{minipage}{\linewidth} 
Make sure that the \inlinecode{bounds} command is supplied with the exact same trajectory as in the GIST calculation. 
Any preprocessing steps need to match between the \inlinecode{bounds} command and the \inlinecode{gist} calculation. 
A helper script called \inlinecode{findcentroid.py} for the same purpose is supplied with this tutorial, and can be used as follows:
~
\begin{lstlisting}[style=bash]
python findcentroid.py -i structure.pdb
\end{lstlisting}
\end{minipage}
\paragraph{Run GIST using cpptraj}
GIST is implemented in \software{cpptraj} which requires the topology and the trajectory of the MD simulation to be analyzed.
The command needed to run GIST in cpptraj can be written in an input file (e.g gist.in) for convenience.
The basic command, assuming a grid centered at coordinates (x=\SI{10.5}{\angstrom}, y=\SI{20.1}{\angstrom}, z=\SI{30.1}{\angstrom}) and a grid size of $30\times30\times30$ voxels with \SI{0.5}{\angstrom} spacing, is:

@@

\subsection{Tutorial data}
We will use the crystal structure of streptavidin (PDB code: 1STP \cite{Weber1989-streptavidin-structure}) to start our calculations.
For section \ref{subsec:gist_biotin}, we will use the biotin ligand extracted from the crystal structure.
For the second part in section \ref{subsec:gist_complex}, we will use the whole crystal structure.
For the sake of simplicity, you can download prepared and solvated Amber topologies and structures based on 1STP by running
\begin{lstlisting}[style=bash]
git clone git@github.com:liedllab/gist-tutorial.git
\end{lstlisting}
This will download the tutorial files, including both the manuscript and the examples in the \inlinecode{code} folder.
Additionally, a link to download the original MD trajectory files is provided in the GitHub repository.
\subsection{Running GIST for Biotin}
\label{subsec:gist_biotin}
In this example, we will be running a simple GIST calculation of the biotin solute molecule to get familiar with the usage of GIST in \software{cpptraj}.
\subsubsection{System preparation}
First, extract biotin from the crystal structure and generate a topology and coordinate file compatible with Amber or a simulation package of your choice.
The solute molecule should be solvated with a sufficiently large solvent box. 
For periodic boundary conditions, the size of the solvent box should be large enough for the solvent to behave bulk-like so as to reduce artifacts from solvent-mediated interactions between the periodic images. 
This corresponds to a distance between solute and box boundary of at least \SI{12}{\angstrom} for water \cite{Chen2021,Waibl2022-gist-solvents} though we recommend \SI{15}{\angstrom} or higher.
Carefully check the employed force field, protonation state and partial charges.
%a PDB file containing only biotin. Prepare your structure by adding hydrogens, making sure to choose the correct protonation state. Parametrize biotin with the forcefield of your choice. Solvate the parametrized biotin, making sure to choose a water box of large enough size so that the water on the box edges behaves bulk-like (a wall distance of at least 15\AA is recommended).
A good starting point for the parametrization and solvation of small molecules is the Amber tutorial 4b (\url{http://ambermd.org/tutorials/basic/tutorial4b/}).

@@

To decide on a grid size, find the extent (minimum and maximum) of the x-, \mbox{y-,} and z-coordinates of the molecule.
To get bulk-like water properties at the grid border, a buffer distance of \SI{15}{\angstrom} is reasonable for biotin.
The grid parameters can be directly calculated using \software{cpptraj's} \inlinecode{bound} command:
\begin{lstlisting}[style=cpptraj]
parm solvated.parm7
trajin md-01.nc 1 1 1
autoimage !(:WAT) origin
bounds !(:WAT) dx 0.5 offset 30 name Grid out bounds.dat
run
\end{lstlisting}
In the above snippet, we calculate the bounds around the whole protein for a grid with \SI{0.5}{\angstrom} side-length voxels and a buffer zone of \SI{15.0}{\angstrom}. 
The atom mask !(:WAT) specifies that the grid should be calculated around all atoms that are not water residues.
Note that the offset is given in terms of number of additional voxels, not length dimension. 
By varying the atom mask in the \inlinecode{bound} command, it is possible to center the grid on a region of interest, similar to the \software{findcentroid.py} script.
